A multi-species benchmark for training and validating mass spectrometry proteomics machine learning models

### 2024-08-20 doc/nine-species/response.tex

\documentclass{article}
\usepackage[margin=1in]{geometry} % Set margins.
\usepackage{url} % Allow URLs.
\usepackage{graphicx} % Allow figures
\usepackage{amsmath} % Allow fancy math.
\usepackage{amssymb}
\usepackage{algorithm, algpseudocode} % Format algorithms
\usepackage[sort&compress]{natbib} % Better bibliography formatting.
\usepackage{xcolor} % Allow colored text.
\usepackage{siunitx}

% Turn off indenting.
\setlength{\parindent}{0pt}
\setlength\parskip{0.5em plus 0.1em minus 0.2em}

\newcommand{\response}[1]{\vspace*{1ex} \color{blue} \noindent #1 \color{black}
	\vspace*{2ex}}

\newcommand{\edit}[1]{\begin{quotation}\color{red}\noindent #1
		\color{black}\end{quotation}}

\newcommand{\myinput}[1]{\begin{quotation}\noindent\color{red}\input{#1}\color{black}\end{quotation}
\vspace*{1ex}
}

\usepackage{soul}
\newcommand{\fixme}[1]{\hl{#1}}

\begin{document}

\noindent
\today\\[2ex]

\noindent
Dear Dr.\ Guo:\\[2ex]

We thank you and the reviewers for your kind consideration of our manuscript.
Below, we address each of the points raised by the two reviewers and describe the changes we have made to the manuscript.
In what follows, the reviewer's comments are shown in black type, interleaved with our responses (in blue) and, where appropriate, the modified text (in red).
We have also made the formatting changes that you requested.

Thank you very much for your consideration.\\[2ex]

\noindent
Best regards,\\[2ex]

\noindent
William Noble\\
University of Washington

\clearpage
\section*{Reviewer 1}

The author pointed out a significant issue in de novo peptide sequencing. I really appreciate his efforts to provide this curated benchmark dataset for the proteomics community. This kind of benchmark dataset is valuable and useful for training high-quality models. However, I have several technical concerns. Here they are:

(1) The author reanalyzed the nine-species dataset using Tide search engine. I highly suggest at least two search engines should be used to make sure the PSMs are correct, since almost every search engine has its own preference.

\response{We agree with you that the choice of search engine introduces a potential bias into the dataset construction.
  However, mitigating this bias is challenging.
  If we combine the results of two different search engine, using either a union or an intersection operation, then we sacrifice the ability to rigorously control the FDR among the reported list of discoveries.}

(2) Why did the author only consider the FDR at PSM level? I think the FDR at protein level is commonly used now.

\response{We agree that, in many settings, controlling the FDR at the protein level is appropriate, especially when a study is aimed at characterizing the protein content of a given set of biological samples.
  However, for constructing a benchmark of PSMs used for training and validating \textit{de novo} sequencing methods, we believe that protein level FDR control is not as relevant.
  In particular, even if we control the FDR at the protein level, the benchmark needs to report a set of PSMs rather than proteins, and it is not clear how to go from a set of detected proteins to a corresponding collection of confident PSMs.
  Consider, for example, the widely used ``picked protein'' method for protein-level FDR control \cite{savitski2015scalable}.
  In the Savitski \textit{et al.}\ (2015) description of this method, each protein is assigned a score equal to the sum of the search engine scores assigned to its peptides.
  With such an approach, a protein can be detected with high confidence even if one or more of its associated peptides has low scores, as long as some of the peptides have very high scores.
  Thus, if we include all PSMs associated with a detected peptide in our benchmark, we risk contaminating the benchmark with incorrect PSMs.}

(3) The annotation in MGF is adding the peptide sequence. Is it possible to provide the annotated product ions? For example, which peak is annotated as b5 ion or y3 ion. With these annotations at ion level, this dataset will be more useful.

\response{Yes, we have added to the benchmark a \fixme{what format?} file that provides this annotation.  Here is the corresponding description from the revised manuscript:}

\fixme{[Bill] Write this.}

\clearpage
\section*{Reviewer 2}

This manuscript describes a new dataset designed for evaluating de novo peptide sequencing algorithms. This new dataset represents an enhancement of the widely-used nine-species dataset, addressing two key issues present in the original dataset:

1. The nine-species dataset incorrectly assigns deamidation modifications to isotopic peaks too frequently.

2. In the nine-species dataset, some peptides are shared across species, which can lead to overly optimistic evaluation outcomes for complex models.

Additionally, a balanced dataset is included for specialized evaluation purposes.

Overall, the dataset generation process is clearly outlined, but some important statistical numbers reported about the generation process need clarification. Specifically, on page 4, in the section titled ``Database Search and FDR Control," the manuscript states that ``At this point... the dataset contains 179,256 distinct peptides." Then, in the next section, it mentions ``Among the 229,984 unique peptides," and after further processing, ``The final, non-redundant benchmark dataset (`main') consists of... 258,105 distinct peptides." It is unclear how these three numbers of peptides are related, and what distinguishes ``unique peptides" from ``distinct peptides." Clarifying these relationships and definitions is essential for assessing the quality of the dataset generation process.

\response{We apologize for the confusion, and we thank the reviewer for pointing out the inconsistency in the reported numbers of peptides in the benchmark.
  We have fixed a bug in the way that our scripts handle PTM annotations in sequences, and the tables in the manuscript as well as the Zenodo files have been updated accordingly.
  \fixme{[Bill] Finish making these changes.}
  We previously used the term ``unique'' and ``distinct'' interchangeably, and in the revised version of the manuscript we stick to ``distinct.''
}

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2024-08-21 doc/nine-species/response.tex

2. In the nine-species dataset, some peptides are shared across species, which can lead to overly optimistic evaluation outcomes for complex models.

Additionally, a balanced dataset is included for specialized evaluation purposes.

Overall, the dataset generation process is clearly outlined, but some important statistical numbers reported about the generation process need clarification. Specifically, on page 4, in the section titled ``Database Search and FDR Control," the manuscript states that ``At this point... the dataset contains 179,256 distinct peptides." Then, in the next section, it mentions ``Among the 229,984 unique peptides," and after further processing, ``The final, non-redundant benchmark dataset (`main') consists of... 258,105 distinct peptides." It is unclear how these three numbers of peptides are related, and what distinguishes ``unique peptides" from ``distinct peptides." Clarifying these relationships and definitions is essential for assessing the quality of the dataset generation process.

\response{We apologize for the confusion, and we thank the reviewer for pointing out the inconsistency in the reported numbers of peptides in the benchmark.
  We have fixed a bug in the way that our scripts handle PTM annotations in sequences, and the tables in the manuscript as well as the Zenodo files have been updated accordingly.
  In the new version, there are 168,422 distinct peptides after the initial construction of the benchmark.
  This number remains the same after eliminating overlaps (since each distinct peptide is randomly assigned to a single species, as described in the text).
  The downsampled benchmark contains 133,232 distinct peptides.

  We previously used the term ``unique'' and ``distinct'' interchangeably, and in the revised version of the manuscript we stick to ``distinct.''
}

### 2024-08-21 doc/nine-species/nine-species.tex
Add draft of reviewer response.  Modify summarization script to list distinct peptides
Update benchmark numbers

\author[1,2]{William Stafford Noble}

\affil[1]{Department of Genome Sciences, University of Washington}
\affil[2]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington}

@@

%Instructions at
%  \url{https://www.nature.com/sdata/publish/submission-guidelines}.}

@@

\begin{table}
  \centering
  \scriptsize
\begin{tabular}{lllrrrrrrrrrr}
\hline
PRIDE & Species & Uniprot & Files & Spectra &
\multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} & Pre & Frag \\
& & & & & PSM & Pep  & PSM & Pep  \\
\hline
PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
108,756 & 13,768 & 102,576 & 13,656 & 20 & 0.05 \\
PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
25,490 & 6833 & 25,490 & 6833 & 10 & 0.05 \\
PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
267,269 & 23,448 & 100,517 & 16,774 & 10 & 0.05 \\
 PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
1,357,908 & 64,025 & 113,613 & 29,245 & 30 & 0.05 \\
PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
82,204 & 10,863 & 82,204 & 10,863 & 20 & 0.05 \\
PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
177,941 & 62,294 & 100,325 & 44,052 & 15 & 0.05 \\
PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
586,278 & 33,008 & 109,023 & 20,057 & 20 & 0.05 \\
PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
193,437 & 30,013 & 101,700 & 24,905 & 20 & 0.05 \\
PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
44,348 & 13,853 & 44,348 & 13,853 & 20 & 0.02 \\
\hline
Total & & & 343 & 15,165,161  & 2,843,631 & 258,105 & 779,796 & 180,238  \\
\end{tabular}
\caption{{\bf Two versions of the nine-species benchmark.}
  Counts of the number of annotated spectra and distinct peptide sequences are provided for the main and balanced versions of the benchmark.
The final two columns specify the precursor window size (in ppm) and fragment bin size (in Da) used in the database search step.
}
\label{tab:benchmark}
\end{table}

@@

\begin{sloppypar}
We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
Tide automatically added to each index a shuffled decoy peptide corresponding to each target peptide.
Thereafter, each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Table~\ref{tab:benchmark}).
The search engine employed XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
For the benchmark, we retained all PSMs with Percolator q value $<$ 0.01.
We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 179,256 distinct peptides.
\end{sloppypar}

@@

To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same, and we converted all isoleucines to leucines.
Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,105 distinct peptides.

@@

At this stage, the benchmark was quite imbalanced, in the sense that some species had a much larger number of associated PSMs.
We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,796 PSMs from 180,238 distinct peptides.

>>>>

\author[1]{Wout Bittremieux}
\author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}

\affil[1]{Department of Computer Science, University of Antwerp, Belgium}
\affil[2]{Department of Genome Sciences, University of Washington, Seattle, WA, USA}
\affil[3]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington, Seattle, WA, USA}

@@

@@

\begin{table}
  \centering
  \scriptsize
\begin{tabular}{lllrrrrrrrrrr}
\hline
PRIDE & Species & Uniprot & Files & Spectra &
\multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} & Pre & Frag \\
& & & & & PSM & Pep  & PSM & Pep  \\
\hline
PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
108,402 & 11,638 & 102,255 & 11,557
& 20 & 0.05 \\
PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
25,522 & 5630 & 25,522 & 5630
& 10 & 0.05 \\
PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
267,183 & 15,220 & 100,485 & 11,934
& 10 & 0.05 \\
PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
1,351,938 & 28,364 & 113,234 & 17,481
& 30 & 0.05 \\
PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
82,514 & 8080 & 82,514 & 8080
& 20 & 0.05 \\
PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
177,553 & 48,459 & 100,056 & 35,787
& 15 & 0.05 \\
PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
585,846 & 19,102 & 108,973 & 13,285
& 20 & 0.05 \\
PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
194,604 & 21,081 & 102,285 & 18,630
& 20 & 0.05 \\
PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
44,555 & 10,848 & 44,555 & 10,848
& 20 & 0.02 \\
\hline
Total & & & 343 & 15,165,161  & 2,838,117 & 168,422 & 779,879 & 133,232 \\
\end{tabular}
\caption{{\bf Two versions of the nine-species benchmark.}
  Counts of the number of annotated spectra and distinct peptide sequences are provided for the main and balanced versions of the benchmark.
The final two columns specify the precursor window size (in ppm) and fragment bin size (in Da) used in the database search step.
}
\label{tab:benchmark}
\end{table}

@@

\begin{sloppypar}
We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
Tide automatically added to each index a shuffled decoy peptide corresponding to each target peptide.
Thereafter, each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Table~\ref{tab:benchmark}).
The search engine employed XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
For the benchmark, we retained all PSMs with Percolator q value $<$ 0.01.
We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 168,422 distinct peptides.
\end{sloppypar}

@@

To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
Among the 168,422 distinct peptides, we identified 4121 (2.4\%) that occur in more than one species.
For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same, and we converted all isoleucines to leucines.
Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
The final, non-redundant benchmark dataset (``main'') consists of 2,838,117 PSMs corresponding to 168,422 distinct peptides.

@@

At this stage, the benchmark was quite imbalanced, in the sense that some species had a much larger number of associated PSMs.
We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,879 PSMs from 133,232 distinct peptides.
